A Diminishing Payment Scheme of Employment Insurance Balances Financial Support and Reemployability for Insurance Recipients^[Code and data are available at; https://github.com/Mikmok14/UI-Spending-and-Income-Implications.git]

### 2022-02-27 outputs/paper/paper.Rmd
Add files via upload

---
title: Unemployment Insurance and Spending
author: "Marco Chau and Chyna Hui"
date: "18 February 2022"
output:
  bookdown::pdf_document2:
header-includes:
  \usepackage{float}
  \floatplacement{figure}{H}
toc = FALSE:
abstract: Abstract goes here

@@

The dataset we use to replicate Ganong and Noel's graphs and tables are downloaded from the American Economic Review's journal website that links to OPENICPSR where the data is stored.^[https://www.openicpsr.org/openicpsr/project/116193/version/V1/view]

@@

## Biases and Weaknesses

@@

@@

Table \@ref(tab:spendchangetypetab) reorganizes table \@ref(tab:spendchangetab) by grouping different spending types together instead. This makes it easier to compare the impact of UI loss on spending between the spending types. Calculating the average percentage loss, spending on durable goods was reduced by 13.14%, nondurable goods reduced by 9%, strict nondurables were reduced by 9.98% and other nondurables were reduced by 15.68%. The largest reduction being other nondurable goods (which includes discount stores, department stores, retail, drug stores, medical copay, and online retail) is within expectations as nonessential goods are the first to be sacrificed during financial duress.

@@

# Discussion
Discussion goes here

# References

>>>>

---
title: A Diminishing Payment Scheme of Employment Insurance Balances Financial Support and Reemployability for Insurance Recipients^[Code and data are available at; https://github.com/Mikmok14/UI-Spending-and-Income-Implications.git]
author: "Marco Chau and Chyna Hui"
date: "18 February 2022"
output:
  bookdown::pdf_document2:
header-includes:
  \usepackage{float}
  \floatplacement{figure}{H}
toc = FALSE:
abstract: This paper replicates the 2019 article *Consumer Spending during Unemployment Positive and Normative Implications*^[@mainrepref DOI:10.1257] which uses income, spending, and employment data to examine the normative implications for UI policy. This replication uses the original data provided by the authors and supplements it with external UI policy theories to evaluate the policy suggestions that the authors make. This replication finds that the UI pool is populated by myopic households that require an extension of financial support which are consistent with the original paper. This replication paper however finds that introducing a payment scheme of diminishing payments in combination with an extension of financial support duration leads to a more efficient market outcome.

@@

The dataset we use to replicate Ganong and Noel's graphs and tables are downloaded from the American Economic Review's journal website that links to OPENICPSR where the data is stored.^[https://www.openicpsr.org/openicpsr/project/116193/version/V1/view] This data was cleansed using the `R programming language`^[@Rref] and packages for R including: `tidyverse`^[@tidyverseref], `ggplot2`^[@ggplotref], `kableExtra`^[@kableExtraref], `dplyr`^[@dplyrref], and `knitr`.^[@knitrref]

@@

## Biases and Limitations

@@

Another limitation of this dataset is that employment status is inferred using direct deposit data. Direct deposit data may lead to false positives when employers change transaction descriptions on pay cheques. False negatives may arise since households can be paid using paper cheques or cash which are not categorized by the JPMCI system.^[@mainrepref page 2405]

@@

Table \@ref(tab:spendchangetypetab) reorganizes table \@ref(tab:spendchangetab) by grouping different spending types together instead. This makes it easier to compare the impact of UI loss on spending between the spending types. Calculating the average percentage loss, spending on durable goods was reduced by 13.14%, nondurable goods reduced by 9%, strict nondurables were reduced by 9.98% and other nondurables were reduced by 15.68%. The largest reduction being other nondurable goods (which includes discount stores, department stores, retail, drug stores, medical copay, and online retail) is within expectations as nonessential goods are the first to be sacrificed during financial duress. The large drops in spending of durable goods at 13.14% however does suggest that during unemployment, household spending becomes myopic and present-focused. Households are more concerned with spending money on immediately consumable goods rather than investments on goods that produce utility over time. The data from this reorganized table supports Ganong and Noel's behavioural model of consumers.^[@mainrepref page 2412]

@@

Figure \@ref(fig:iandsfig) is the result of plotting both the income (figure \@ref(fig:incomefig)) and spending (figure \@ref(fig:spendingfig)) data onto the same plane. The solid black line tracks income while the grey line tracks spending. Note that the y-axis denotes the ratio of income and spending relative to y at x = -5 and do not represent absolute dollar values of income and spending. The vertical blue line at x = 0.5 marks the approximate start of when the first UI payment is deposited into the household's bank account. The vertical red line at x = 5.5 marks approximately when the last UI payment is deposited. The horizontal purple line at y = 1 serves as a marker to original income and spending 5 months before the first UI payment, which we assume to be when the recipient was still employed.

Juxtaposing the income and spending data, we both income and spending drop at similar times, however the degree of the reductions are quite different. In general, we observe that income drops much more than spending.

Both income and spending experience an initial drop 3 months before the the first UI payment (x = -3) and both decrease rapidly faster 2 months before the first UI payment (x = -2). At the start of receiving UI, spending is at around 92% of employment levels what income is only at 80% meaning income has dropped 12% more than spending. The simultaneous drop of both income and spending suggests that there is a positive relationship between these two variables, however the discrepancy in their rates of change indicates that this is not a 1:1 relationship. Once UI deposits start, spending and income both become much more level, though they both exhibit a slight decrease from months 0 - 3.

After the last month of UI income, both spending and income drop dramatically. Spending drops from 90% at 5 months to 78% at 6 months, a drop of 12% over 1 month. Income on the other hand drops from 75% to 32%, a decrease of 43% in the same time frame. At 6 months since the first UI payment, income is at 32% of pre-unemployment levels while spending is at 78%, a discrepancy of about 46% compared to the 12% at the beginning of UI deposits.

# Discussion

Figure \@ref(fig:iandsfig) supports Ganong and Noel's hypothesis that a myopic consumer model reflects unemployed consumption patterns more so than the standard rational model.^[@mainrepref page 2407]. This is evident in the discrepancy between income and spending while the household was receiving UI versus after. To smoothen consumption, a rational model of forward-looking agents would reduce their spending to an equilibrium level without accounting for UI income. This is because UI income is temporary and lasts only for 6 months. The sudden drop in spending after UI has been exhausted suggests that households are present-minded and reactionary to the loss of income. Since spending drops from 90% while receiving UI to 78% after UI has been exhausted relative to spending while employed, this suggests that households were spending 12% above their equilibrium level when not accounting for UI income. A perfectly rational forward-looking agent would predict the loss of income after 6 months and save that extra 12% that they are spending to be used at a later time when marginal utility of each dollar is greater than at the present. Thus we would expect that spending does not change as the household deposits their last UI cheque, which is not reflected in the data. However, the large drop in spending at this period suggests that UI recipients are present-focused and are not basing their spending decisions on long-term gains which is also supported by Carbone and Hey's field analysis of consumption behaviours of unemployed consumers.^[@reactref] Furthermore, the reduction in household spending on durable goods which provide long term utility by 13.14% compared to strict nondurable goods and nondurable goods by 9.98% and 9% respectively shown in table \@ref(tab:spendchangetypetab) shows a shift in spending preference for immediately consumable goods rather than goods that payoff over time; a shift from future-focus to present-focused spending. This finding results in the implication that UI policies need to be designed to best benefit present-focused myopic agents to optimize market outcomes.

The large difference between income and spending after UI exhaustion at 6 months is concerning from a policy perspective because of its potential impact on reemployability. As shown in figure \@ref(fig:employfig), employment hazard drops substantially right after household stop receiving UI income. A potential source of this drop in employment hazard is the drop in life satisfaction discussed by Rose and Stavrova.^[@satisref] Since spending and income decline in tandem quite slowly during the period where the household is receiving UI, we postulate that the dollar values of spending and income during this time are roughly equivalent and is at an equilibrium; households are neither saving money nor spending their savings when the difference between income and spending is 15%.^[This number is derived by subtracting the y-values of income and spending at x = 5]  Once UI ends however, the difference enlarges to 46%. Despite both spending and income dropping, the larger drop in income relative to spending suggests that households are under more financial pressure than before the UI income was exhausted. An increase in financial duress could lead to lower perception of life-satisfaction which has a negative impact on reemployability probability.^[@satisref] Ganong and Noel's suggestion to extend the duration of UI benefits^[@mainrepref page 2417] is able to alleviate financial duress caused by lowered income and thereby promoting reemployment, however Rose and Stavrova's study also shows that excessive life satisfaction also discourages reemployment. The parabolic relationship of life satisfaction and reemployment suggests that UI policies need to strike a balance between alleviating financial pressure for households to help them stay afloat, but also apply enough pressure to UI recipients to continue job-seeking before UI exhaustion to maximize reemployability.

Combining the above two policy implications, we suggest a UI schedule that deposits benefits over a period of about 9 - 12 months that diminishes monthly. By extending UI benefits for 3 - 6 more months, this schedule takes advantage of the increase in marginal benefit of UI income over time compared to a dollar increase to the existing payment schedule of 6 months.^[@mainrepref page 2418] Extension of benefit duration also extends the higher hazard rate of employment shown in figure \@ref(fig:employfig) where employment hazard has a much lower equilibrium level after UI exhaustion than before. Diminishing each subsequent UI payment so that it is lower than the previous payment serves to both reduce the sudden drops in spending after UI is exhausted, and to create mounting pressure on recipients to find employment.^[@satisref] Using the myopic model of unemployed consumers, a diminishing payment scheme weans households off of the UI so that their spending falls in tandem with income gradually. This payment scheme also discourages recipients from intentionally avoiding employment to collect UI benefits. The use of this extended and diminishing payment scheme aims to support myopic households to adjust their spending to reflect their new unemployed income and encourage them to find employment sooner than the month UI ends.

## Next Steps

The literature used to support the arguments made in this paper provide evidence from both American and European UI samples. The differences in both payment schedule, payment amount, payment duration, and labour market between these economies may lead to discrepancy between assumptions and conclusions that can be made in one market and not the other. In order to understand possible outcomes of changes to American UI policy more consistently, more American data and samples should be gathered and used.

As mentioned in section \@ref(biases-and-limitations), the use of only JPMCI data limits the findings in this report to reflect JPMCI customers. To understand the UI landscape more holistically, financial data from the other big four banks, and smaller alternative banks should be examined as well. Households of other banks may have access to financial services and tools that impact their spending choices during unemployment. Analyzing other bank data also allows us to see how households with multiple bank accounts distribute their liquid assets during unemployment and UI benefits which is obscured to us in this dataset. Spending and income may be higher than the JPMCI data suggests.

Section \@ref(biases-and-limitations) also discusses the limitation of only using banking data to infer employment status. To limit false positives and false negatives due to limitations in the JPMCI system, a study where employment status for UI recipients is gathered using census data or direct interview/survey data would be preferred. This also creates a more accurate picture of when UI recipients were reemployed to avoid clerical discrepancies caused by cheques.

\newpage
# References
